---
title: "p8105_hw3_td2749"
author: "Tvisha R. Devavarapu"
date: "2022-10-07"
output: github_document
---

```{r Setting up tidyverse}
library(tidyverse)
```


## Problem 2: Accelerometer Data

* Load, tidy, and otherwise wrangle the data. Your final dataset should include all originally observed variables and values; have useful variable names; include a weekday vs weekend variable; and encode data with reasonable variable classes. Describe the resulting dataset (e.g. what variables exist, how many observations, etc).

```{r}
accel_data = 
  read_csv("./data/accel_data.csv") %>%
  janitor::clean_names() %>%
  mutate(day_type = case_when(
    day == "Saturday" ~ "weekend",
    day == "Sunday" ~ "weekend",
    day == "Monday" ~ "weekday",
    day == "Tuesday" ~ "weekday",
    day == "Wednesday" ~ "weekday",
    day == "Thursday" ~ "weekday",
    day == "Friday" ~ "weekday")) %>% 
  select(week, day_id, day, day_type, everything())

accel_data = mutate(accel_data, total_activity = rowSums(accel_data[5:1444]))
accel_data = select(accel_data, week, day_id, day, day_type, total_activity, everything())
```


* Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each day, and create a table showing these totals. Are any trends apparent?

* Accelerometer data allows the inspection activity over the course of the day. Make a single-panel plot that shows the 24-hour activity time courses for each day and use color to indicate day of the week. Describe in words any patterns or conclusions you can make based on this graph.
